Biology background

### 2015-04-21 Rmarkdown/biology_background.Rmd
Changes suggested by Belinda to Biology Background

The nucleosomes are interconnected by _linker DNA_ to form the macromolecule called chromatin. The linker sequences are between $20$ to $60$ nucleotides in length, while approximately 147 nucleotides are wrapped around each nucleosome and a further 20 nucleotides wrapped around each additional H1/H5 histone \citep{Annunziato:2008ue}.

@@

As is clear from the above description, the genome is a complex three-dimensional structure. Nonetheless, in bioinformatics and computational biology, the genome is mostly considered as a single-stranded, one-dimensional string of the nucleobases _A_, _C_, _G_ and _T_.

@@

Variation between individual's genomes, be it at single nucleotides[^snvs_snps] or across larger regions[^indels], is one important source of phenotypic variation. Importantly, genetically-driven phenotypic variation is frequently _heritable_, meaning that phenotypes can be passed on from one generation to the generation via the genome.

@@

When people speak of DNA methylation they are generally referring to methylation of the cytosine nucleobase. Even more specifically, they are referring to the molecule _5-methylcytosine_. 5-methylcytosine, abbreviated as _5mC_, is by far the most common form of DNA methylation in the animal and plant kingdoms[^other_DNAm]. However, I will continue to use the term _DNA methylation_ to describe the more specific 5mC, as is standard in the literature.

@@

A German chemist, W.G. Ruppel, first identified a methylated nucleic acid in 1898. Ruppel was studying _tuberculinic acid_, the poison of _Mycobacterium tuberculosis_[^tuberculosis] , and discovered that it contained a methylated nucleotide \citep{Ruppel:1899hc}. In 1925, Johnson and Coghill isolated 5-methylcytosine as a product of hydrolysis of tuberculinic acid, the nucleic acid of _Mycobacterium tuberculosis_ \citep{Johnson:1925js}. However, Johnson and Coghill's results were disputed for over twenty years by other researchers who were unable to replicate the original findings \citep{VISCHER:1949ty}.

@@

The typical site of DNA methylation is at the C5 carbon position of a cytosine nucleotide. This modified cytosine is called _5-Methylcytosine_ or _5mC_. Figure \ref{fig:5mC} shows the structure of 5mC.

@@

The importance of 5-methylcytosine in mammalian genomes is such it has been dubbed the "fifth base" of the DNA code \citep{Lister:2009boa}. In mammalian genomes, most cytosines are unmethylated except for those at _CpG dinucleotides_. A _CpG dinucleotide_, or more simply a _CpG_, is a cytosine followed by a guanine in the linear DNA sequence. The 'p' stands for the phosphate backbone of DNA and some authors omit it in favour of simply calling it CG methylation. A CpG is a palindromic sequence and is generally symmetrically methylated. Approximately 70% of CpGs are methylated in mammals \citep{Laird:2003kw}, meaning that the cytosine in the CpG dinucleotide is a 5-methylcytosine.

@@

CpGs are underrepresented in the human genome. The _GC-content_ of the human genome, which is defined as the percentage of nucleotides that are either guanines or cytosines, is approximately $41\%$. If these nucleotides were uniformly distributed across the genome then we would expect about $4.1\%$ of _dinucleotides_, pairs of nucleotides, to be CpGs. Instead, only $1\%$ of dinucleotides are CpGs.

One reason for the relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines \citep{Scarano:1967ue}. Thus, over time, many methylated CpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see Figure \ref{fig:dinucleotide_frequency_hg19}). There are many other evolutionary pressures on the distribution of nucleotides in a genome. One effect of this is that the distribution of CpGs is far from uniform. In fact, CpGs tend to form clusters, which are termed CpG islands.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/dinucleotide_frequency_hg19.pdf}
\caption{Observed:expected ratio of dinucleotides in the human reference genome (hg19). The expected frequency is computed under an 'independence' model, based on the observed frequencies of each nucleotide.}
\label{fig:dinucleotide_frequency_hg19}
\end{figure}

@@

CpG islands contain the 20-40% of CpGs that are frequently unmethylated in mammalian genomes. CpG islands are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found \citep{Wu:2010do}. The classical definition of a CpG island, given by \citet{GardinerGarden:1987we}, and used by the popular UCSC genome browser ([http://genome.ucsc.edu/cgi-bin/hgGateway](http://genome.ucsc.edu/cgi-bin/hgGateway)), is a region of the genome where:

@@

In humans, cytosine methylation in most cell types is found almost exclusively at CpGs \citep{Jones:2012jr}. There are, however, certain cell types with widespread non-CpG methylation. Non-CpG methylation is often classified as CHH methylation or CHG methylation, where H is the IUPAC code for any nucleotide except G ([http://www.bioinformatics.org/sms2/iupac.html](http://www.bioinformatics.org/sms2/iupac.html)). The rule-of-thumb for mammalian genomes is that non-CpG methylation is rare in somatic cells but common in pluripotent cells. Of course, there are exceptions to every rule, especially in biology.

@@

Overall, non-CpG methylation in humans is less well studied and less well understood than CpG methylation. This is partly due to sampling bias since commonly used assays for studying DNA methylation, such as the Illumina 27k and 450k microarrays, only measure CpG methylation. However, recent technological advances mean that cytosine methylation can be routinely assayed regardless of the sequence context (see Section \ref{sec:dna_methylation_assays}).

@@

The choice of which assay to use for an experiment is a trade-off between resolution, per-sample throughput and per-cost throughput. Experiments that use an assay with high-resolution and high per-sample throughput generally have fewer samples than experiments using a lower resolution assay or an assay with lower per-sample throughput.

In this section I will describe each of the pre-treatments but will focus on the bisulfite-conversion assays. In particular, I will describe in detail the 'gold standard' assay of DNA methylation, _whole-genome bisulfite-sequencing_, that combines the sodium bisulfite conversion pre-treatment with massively parallel sequencing to produce whole-genome maps of DNA methylation at base pair resolution.

@@

While restriction enzyme assays do not typically provide single nucleotide resolution data, `methylCRF` \citep{Stevens:2013hv} is able to infer single CpG methylation levels by combining data from a restriction enzyme based assay (MRE-seq) with one based on affinity enrichment (MeDIP-seq) in a sophisticated statistical analysis.

@@

Bisulfite-conversion assays are considered the gold standard for studying DNA methylation since cytosine methylation can be detected at single-base resolution \citep{Stirzaker:2014cu}. In fact, single-molecule, base pair resolution is even possible for short DNA sequences when bisulfite-treated DNA is analysed with sequencing[^microarrays].

[^microarrays]: Microarray hybridisation assays can provide base pair resolution but not single molecule resolution. The signal from a microarray-based experiment a sample-wide average since, for each locus, the signal is relative to the proportion of DNA fragments in the sample methylated at that locus.

Almost all bisulfite-based assays require a considerable amount of DNA, e.g., $1$ to $5$ $\mu$g for whole-genome bisulfite-sequencing, which is extracted from a population of cells. However, the minimal amount of DNA is being reduced with each technological advance. Recently, both \citet{Guo:2013ih} and \citet{Smallwood:2014kn} published single-cell bisulfite-sequencing assays, although they are not yet a commercially available sequencing assay. Furthermore, these are not yet proper genome-wide assays. The technique of \citet{Guo:2013ih} is adapted from RRBS and so only assays a small percentage of cytosines in the genome. And the technique of \citet{Smallwood:2014kn}, while intended as a genome-wide assay, can reportedly only "accurately measure DNA methylation at up to 48.4% of cpG sites" \citep{Smallwood:2014kn}.

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/bisulfite_CT_SNP.pdf}
\caption{Sequence variation between the reference genome and the sample's genome can result in incorrect inferrence about the methylation state of the sample's genome. The locus in orange is a cytosine in the refererence genome but a thymine in the sample's genome. Because the read is compared against the reference genome, it may incorrectly inferred to be an unmethylated cytosine. The locus in purple is a thymine in the reference genome but an unmethylated cytosine in the sample's genome. Because the read is compared against the reference genome it may incorrectly inferred to be a thymine.}
\label{fig:bisulfite_CT_SNP.pdf}
\end{figure}

@@

The only commercially available assay in this class is the Pacific Biosciences SMRT technology \citep{Flusberg:2010hra}. It infers the presence of DNA modifications by comparing the time it takes to 'read' the modified form of a nucleotide, such as 5mC, to the time it takes to read its unmodified form.

@@

Whole-genome bisulfite-sequencing provides single-nucleotide resolution data and is the gold standard assay for studying DNA methylation. Methods for analysing whole-genome bisulfite-sequencing data are the focus of my thesis. The following chapter describes the bioinformatics analysis of a bisulfite-sequencing experiment, including its many statistical and computational challenges.

>>>>

The nucleosomes are interconnected by _linker DNA_ to form the macromolecule called chromatin. The linker sequences are between $20$ to $60$ basepairs of DNA (bp) in length, while approximately $147$ bp are wrapped around each nucleosome and a further $20$ bp wrapped around each additional H1/H5 histone \citep{Annunziato:2008ue}.

@@

As is clear from the above description, the genome is a complex three-dimensional structure. Nonetheless, in bioinformatics and computational biology, the genome is mostly considered as a single-stranded, one-dimensional string of the bases _A_, _C_, _G_ and _T_.

@@

Variation between individual's genomes, be it at single base[^snvs_snps] or across larger regions[^indels], is one important source of phenotypic variation. Importantly, genetically-driven phenotypic variation is frequently _heritable_, meaning that phenotypes can be passed on from one generation to the generation via the genome.

@@

When people speak of DNA methylation they are generally referring to methylation of the cytosine base. Even more specifically, they are referring to the molecule _5-methylcytosine_. 5-methylcytosine, abbreviated as _5mC_, is by far the most common form of DNA methylation in the animal and plant kingdoms[^other_DNAm]. However, I will continue to use the term _DNA methylation_ to describe the more specific 5mC, as is standard in the literature.

@@

A German chemist, W.G. Ruppel, first identified a methylated nucleic acid in 1898. Ruppel was studying _tuberculinic acid_, the poison of _Mycobacterium tuberculosis_[^tuberculosis] , and discovered that it contained a methylated base \citep{Ruppel:1899hc}. In 1925, Johnson and Coghill isolated 5-methylcytosine as a product of hydrolysis of tuberculinic acid, the nucleic acid of _Mycobacterium tuberculosis_ \citep{Johnson:1925js}. However, Johnson and Coghill's results were disputed for over twenty years by other researchers who were unable to replicate the original findings \citep{VISCHER:1949ty}.

@@

The typical site of DNA methylation is at the C5 carbon position of a cytosine base. This modified cytosine is called _5-Methylcytosine_ or _5mC_. Figure \ref{fig:5mC} shows the structure of 5mC.

@@

The importance of 5-methylcytosine in mammalian genomes is such it has been dubbed the "fifth base" of the DNA code \citep{Lister:2009boa}. In mammalian genomes, most cytosines are unmethylated except for those at _CpG dinucleotides_. A CpG dinucleotide, or more simply a _CpG_, is a cytosine followed by a guanine in the linear DNA sequence. The 'p' stands for the phosphate backbone of DNA and some authors omit it in favour of simply calling it CG methylation. A CpG is a palindromic sequence and is generally symmetrically methylated. Approximately $70\%$ of CpGs are methylated in mammals \citep{Laird:2003kw}, meaning that the cytosine in the CpG dinucleotide is a 5-methylcytosine.

@@

CpGs are underrepresented in the human genome. The _GC-content_ of the human genome, which is defined as the percentage of bases that are either guanines or cytosines \citep{Benjamini:2012er}, is approximately $41\%$. If these bases were uniformly distributed across the genome then we would expect about $4.1\%$ of dinucleotides to be CpGs. Instead, only $1\%$ of dinucleotides are CpGs.

One reason for the relative scarcity of CpGs is that methylated cytosines can spontaneously deaminate to thymines \citep{Scarano:1967ue}. Thus, over time, many methylated CpGs will become TpGs, leading to a genome-wide reduction in the proportion of CpGs and a genome-wide increase in the proportion of TpGs (see Figure \ref{fig:dinucleotide_frequency_hg19}). There are many other evolutionary pressures on the distribution of bases in a genome. One effect of this is that the distribution of CpGs is far from uniform. In fact, CpGs tend to form clusters, which are termed CpG islands.

\begin{figure}[!htb]
\centering
\includegraphics[width=\textwidth]{../figures/dinucleotide_frequency_hg19.pdf}
\caption{Observed:expected ratio of dinucleotides in the human reference genome (hg19). The expected frequency is computed under an 'independence' model, based on the observed frequencies of each base.}
\label{fig:dinucleotide_frequency_hg19}
\end{figure}

@@

CpG islands contain the $20\%$ to $40\%$ of CpGs that are frequently unmethylated in mammalian genomes. CpG islands are important regulatory elements in the genome and are where most differences in DNA methylation between different cell types are found \citep{Wu:2010do}. The classical definition of a CpG island, given by \citet{GardinerGarden:1987we}, and used by the popular UCSC genome browser ([http://genome.ucsc.edu/cgi-bin/hgGateway](http://genome.ucsc.edu/cgi-bin/hgGateway)), is a region of the genome where:

@@

In humans, cytosine methylation in most cell types is found almost exclusively at CpGs \citep{Jones:2012jr}. There are, however, certain cell types with widespread non-CpG methylation. Non-CpG methylation is often classified as CHH methylation or CHG methylation, where H is the IUPAC code for any base except G ([http://www.bioinformatics.org/sms2/iupac.html](http://www.bioinformatics.org/sms2/iupac.html)). The rule-of-thumb for mammalian genomes is that non-CpG methylation is rare in somatic cells but common in pluripotent cells. Of course, there are exceptions to every rule, especially in biology.

@@

Overall, non-CpG methylation in humans is less well studied and less well understood than CpG methylation. This is partly due to sampling bias since commonly used assays for studying DNA methylation, such as the Illumina 27k and 450k microarrays, measure almost exclusively CpG methylation. However, recent technological advances mean that cytosine methylation can be routinely assayed regardless of the sequence context (see Section \ref{sec:dna_methylation_assays}).

@@

The choice of which assay to use for an experiment is a trade-off between resolution, per-sample throughput and per-cost throughput. Experiments that use an assay with high resolution and high per-sample throughput generally have fewer samples due to the associated higher costs than experiments using a lower resolution assay or an assay with lower per-sample throughput.

In this section I will describe each of the pre-treatments but will focus on the bisulfite-conversion assays. In particular, I will describe in detail the 'gold standard' assay of DNA methylation, _whole-genome bisulfite-sequencing_, that combines the sodium bisulfite conversion pre-treatment with massively parallel sequencing to produce whole-genome maps of DNA methylation at single-base resolution.

@@

While restriction enzyme assays do not typically provide single-base resolution data, the `methylCRF` software \citep{Stevens:2013hv} is able to infer single CpG methylation levels by combining data from a restriction enzyme based assay (MRE-seq) with one based on affinity enrichment (MeDIP-seq) in a sophisticated statistical analysis.

@@

Bisulfite-conversion assays are considered the gold standard for studying DNA methylation since cytosine methylation can be detected at single-base resolution \citep{Stirzaker:2014cu}. In fact, single-molecule, single-base resolution is even possible for short DNA sequences when bisulfite-treated DNA is analysed with sequencing[^microarrays].

[^microarrays]: Microarray hybridisation assays can provide single-base resolution but not single-molecule resolution. The signal from a microarray-based experiment a sample-wide average since, for each locus, the signal is relative to the proportion of DNA fragments in the sample methylated at that locus.

Almost all bisulfite-based assays require a considerable amount of DNA, e.g., $1$ to $5$ $\mu$g for whole-genome bisulfite-sequencing, which is extracted from a population of cells. However, the minimal amount of DNA is being reduced with each technological advance. Recently, both \citet{Guo:2013ih} and \citet{Smallwood:2014kn} published single-cell bisulfite-sequencing assays, although they are not yet a commercially available sequencing assay. Furthermore, these are not yet proper genome-wide assays. The technique of \citet{Guo:2013ih} is adapted from RRBS and so only assays a small percentage of cytosines in the genome. And the technique of \citet{Smallwood:2014kn}, while intended as a genome-wide assay, can reportedly only "accurately measure DNA methylation at up to $48.4\%$ of cpG sites" \citep{Smallwood:2014kn}.

@@

\begin{figure}[h]
\includegraphics[width=\textwidth]{../figures/bisulfite_CT_SNP.pdf}
\caption{Sequence variation between the reference genome and the sample's genome can result in incorrect inference about the methylation state of the sample's genome. The locus in orange is a cytosine in the reference genome but a thymine in the sample's genome. Because the read is compared against the reference genome, it may incorrectly inferred to be an unmethylated cytosine. The locus in purple is a thymine in the reference genome but an unmethylated cytosine in the sample's genome. Because the read is compared against the reference genome it may incorrectly inferred to be a thymine.}
\label{fig:bisulfite_CT_SNP.pdf}
\end{figure}

@@

The only commercially available assay in this class is the Pacific Biosciences SMRT technology \citep{Flusberg:2010hra}. It infers the presence of DNA modifications by comparing the time it takes to 'read' the modified form of a base, such as 5mC, to the time it takes to read its unmodified form.

@@

Whole-genome bisulfite-sequencing provides single-base resolution data and is the gold standard assay for studying DNA methylation. Methods for analysing whole-genome bisulfite-sequencing data are the focus of my thesis. The following chapter describes the bioinformatics analysis of a bisulfite-sequencing experiment, including its many statistical and computational challenges.
